\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-05-26 manuscript_v5/main.tex
working on map

\usepackage{rotfloat}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata},
  pdfauthor={Ariel Mundo Ortiz1,2; Bouchra Nasri1,2,},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Differences in COVID-19 vaccination in the province of
Ontario across Health Regions and socio-economic strata}}
\author{}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, sharp corners, enhanced, frame hidden]}{\end{tcolorbox}}\fi

\textsuperscript{1} Centre de Recherches Mathématiques, University of
Montreal, Montréal, Canada\\
\textsuperscript{2} Department of Social and Preventive Medicine, École
de Santé Publique, University of Montreal, Montréal, Canada

\textsuperscript{*} Correspondence:
\href{mailto:bouchra.nasri@umontreal.ca}{Bouchra Nasri
\textless{}bouchra.nasri@umontreal.ca\textgreater{}}

@@

As of May of 2023 there have been 765 million confirmed cases of
COVID-19 around the world, including 6.8 million
deaths\textsuperscript{\protect\hyperlink{ref-WHO-Covid}{1}}. Although
this disease is no longer categorized as a global health emergency by
the World Health Organization
(WHO)\textsuperscript{\protect\hyperlink{ref-rigby2023}{2}}, there is
ongoing concern due to continued transmission, surges in cases and
deaths due to new
variants\textsuperscript{\protect\hyperlink{ref-un2023}{3}}, and
weaknesses in health systems around the world that could be exploited by
a novel virus or another public health emergency in the
future\textsuperscript{\protect\hyperlink{ref-mackey2021}{4}}.

In particular, a major weakness that has received attention during the
pandemic has been related to inequalities in vaccine uptake. The rapid
development of vaccines against COVID-19 initially brought the hope of a
rapid end to the pandemic due to the start of vaccination campaigns in
certain parts of the world toward the end of
2020\textsuperscript{\protect\hyperlink{ref-thelancet2021}{5}--\protect\hyperlink{ref-tanne2020}{8}})
but inequalities in vaccine uptake made these pharmaceutical
interventions ultimately unable to replicate the experience of smallpox,
where vaccination on a global scale and was crucial to control this
disease\textsuperscript{\protect\hyperlink{ref-kayser2021}{9}}.

This problematic is a multifaceted issue resulting from a combination of
factors, among which are failed public health
measures\textsuperscript{\protect\hyperlink{ref-li2021}{10}}, inequality
in vaccine access between high- and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{11},\protect\hyperlink{ref-tamey2022}{12}},
and vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{13}}.
Furthermore, it is well established that this issue has affected in
particular individuals in certain equity-deserving groups (e.g., Black,
Asian, or Indigenous) as well as individuals with socio-economic
disadvantages\textsuperscript{\protect\hyperlink{ref-willis2021}{14}--\protect\hyperlink{ref-hussain2022}{20}}.

Reasons given for this inequality have included medical mistrust due to
systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{16},\protect\hyperlink{ref-mosby2021}{21}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{14}}, and
the influence of conspiracy
theories\textsuperscript{\protect\hyperlink{ref-mosby2021}{21}--\protect\hyperlink{ref-freeman2020}{23}}.
However, it is important to also consider that vaccination uptake can be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{24}},
spatial differences in vaccine access and supply, vaccination location
availability, and lack of prioritization of areas where vulnerable
groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-nguyen2021}{25}}.
Other studies have also shown heterogeneity in vaccine uptake within
small governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{26}--\protect\hyperlink{ref-bhuiyan2022}{29}},
and that accounting for geographical differences in vaccination can help
predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{30}}.

However, such analyses have been carried mostly in territories outside
of Canada, where available studies have been focused in certain cities
(such as Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{31}},
or Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{32}}),
or have explored differences at a province-wide
level\textsuperscript{\protect\hyperlink{ref-guay2022}{18}}. Therefore,
there is a need for studies that explore spatial differences in
vaccination within the Canadian territory and that consequently, can
help identify disparities that need to be addressed within specific
areas in each province.

This need is specially important in the case of Ontario, the most
populated province of Canada. Between 2007 and 2019, Ontario managed
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integration Networks (LHINs), which aimed to
provide an integrated health system for the province. However, this
approach was complex and bureaucratic, and resulted in excessive
expenditures, disparities in mortality rates, the deterioration of
certain performance indicators such as wait times and hospital
readmissions, fragmented electronic health systems, the decline of
performance indicators, and inequities in health services
access\textsuperscript{\protect\hyperlink{ref-tsasis2012}{33}--\protect\hyperlink{ref-lysyk2016}{37}}.
Therefore, with the intent of better organizing and delivering care in
late 2019 the provincial government eliminated the LHINs and
incorporated the areas covered by them into six larger Health Regions
(North East, North West, Central, Toronto, West, and
East)\textsuperscript{\protect\hyperlink{ref-dong2022}{35}}.

Because the relatively recent adoption of the Health Region model and
its alignment with the onset of the COVID-19 pandemic, there is a need
to analyze if there are ongoing disparities in health access under this
approach that need to be addressed before they are exploited by a new
disease or public health threat. In this regard, previous research has
highlighted disparities in the level of activity of each Health
Region\textsuperscript{\protect\hyperlink{ref-sethuram2023}{38}}.
Therefore, analyzing differences in vaccination uptake within the Health
Regions and can help identify which socio-demographic groups are the
most vulnerable and what areas of the province deserve special attention
by decision-makers.

Therefore, in this study we hypothesized that there were differences in
vaccination uptake between the different Health Regions of Ontario
between October of 2021 and January of 2022. By including socio-economic
factors in our analysis, we aimed at identifying in which groups these
differences were significant in order to provide an assessment of the
current state of healthcare access in Ontario.

@@

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences and the Mathematical Modelling of
COVID-19 Task Force under ethical guidance from the University of
Toronto, and which ran between September 30th, 2021 and January
17th,2022. The survey collected socio-economic information from
participants (Table~\ref{tbl-descriptive-stats}), their location
(nearest municipality, as shown in Figure~\ref{fig-map}), the date of
access to the survey, and asked information on vaccination status by
using the question ``Have you received the first dose of the COVID
vaccine?'', with possible answers ``yes'' and ``no''. The original
dataset contained 39,029 observations.

By design, the survey allowed respondents to exit at any time and
deployed the questions randomly, which resulted in \(\approx\) 84\% of
the observations having multiple missing answers or being incomplete.
Therefore, we selected 6,343 observations that were labeled as
``complete'' in the dataset and that had answers for all covariates
considered in our analysis. Later, we matched the city of each
observation with its corresponding LHIN and Health Region, and removed
observations from areas with low representation (254 observations
corresponding to the North West and North East Health Regions). Finally,
we removed outliers from the data (19 observations of individuals with
household size of 1 and income above 110,000 CAD in the original
dataset). After all the preliminary analyses indicated above, the total
number of observations used for analysis was 6,236 and included the
East, Central, Toronto, and West Health Regions covering the period
between October 1st,2021 and January 17, 2022. The original dataset,
clean dataset, and details on the data cleaning process are described in
detail in the \href{https://github.com/aimundo/Fields_COVID-19/}{GitHub
repository} for this paper.

@@

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{39}--\protect\hyperlink{ref-cnat2022a}{41}}.
Because we identified differences in representativity between the survey
data and the estimates from the Census, we used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{42}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{43}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{44}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{45}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{46}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{47}}.

@@

Figure~\ref{fig-models} presents the estimates (as odd ratios) from the
logistic regression models for vaccination status using the
socio-demographic factors collected by the survey, and their
interactions. Generally speaking, lower odds of vaccination were
identified in both cases in individuals characterized by a low household
income, or that identified as part of equity-deserving groups. However,
the magnitude of the estimates differed between the uncorrected and
corrected models and more importantly, there were differences in the
statistical significance of certain estimates before and after the
correction. Specifically, the uncorrected model showed significant
differences in vaccination odds between the age groups considered, the
East Health Region, Latin American individuals with a household income
under CAD 25,000, and Indigenous individuals living in the Central
Health Region (Figure~\ref{fig-models},B) but these were deemed non
statistically-significant after the correction.

@@

Interestingly, individuals in equity-deserving groups with a household
income below CAD 25,000 had higher odds of vaccination (when compared to
those with a household income above CAD 60,000). This held true in the
case of Arab/Middle Eastern individuals (OR=34, CI={[}1.70,6.79{]}),
Black individuals (OR=3.81, CI={[}2.05, 7.09{]}), and those in other
racial or ethnic groups (OR=3.19, CI={[}2.00,5.09{]}). Additionally,
individuals with an income between CAD 25,000 and 59,999 in the
Arab/Middle Eastern and other racial ethnic groups also had higher odds
of vaccination than their high-income peers (ORs and CIs=6.96
{[}2.67,18.16{]}, and 3.5 {[}1.85,6.62{]}).

Finally, significantly lower odds of vaccination were identified (when
compared to the Toronto Health Region) for Black individuals in the
Central Health Region (OR=0.39, CI={[}0.2,0.75{]}), Arab/Middle Eastern
individuals in the East Health Region (OR=0.41 {[}0.17, 0.98{]}), and in
the Indigenous and mixed groups in the West Health Region (ORs and
CIs={[}0.31 {[}0.14, 0.7{]} and 0.38 {[}0.19, 0.76{]}, respectively).

@@

In this study we hypothesized that differences in COVID-19 vaccination
uptake were present between the Health Regions during between late 2021
and early 2022. Our goal was to determine which socio-demographic groups
could be impacted by these disparities in order to provide
decision-makers with information that could be used to develop policies
focused on reducing or eliminating these differences and ensure that the
Health Region model is able to fulfill its mission of improving health
access for all Ontarians.

@@

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in equity-deserving Canadian individuals to concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{41},\protect\hyperlink{ref-basta2022}{49}--\protect\hyperlink{ref-cnat2023}{51}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of equity-deserving
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{52}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{53}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{54}}. This
suggests that the observed differences are associated to disparities in
vaccine access that were present during the period covered by the
survey.

Interestingly, whereas overall self-reported vaccination rates were
found to be statistically significantly lower in various racial minority
groups when compared to White/Caucasian individuals, the change in odds
of vaccination within certain racial groups and income strata was
actually positive, in contrast to the White/Caucasian group, where
vaccination odds decreased in income brackets below CAD 60,000
(Supplementary Figure A-5). Specifically, individuals in low income
brackets that belonged to Arab/Middle Eastern, Black, or other minority
groups had higher odds of vaccination that their peers with an income
above 60,000 CAD.

This result is likely reflects in part the fact that individuals in
racial minority groups tend to perform occupations that have been deemed
as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{55},\protect\hyperlink{ref-ct2021}{56}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. In Ontario, these workers had priority for
COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{57}}; and
there is evidence of interventions by vaccination staff in certain parts
of the province to encourage vaccination uptake by these
individuals\textsuperscript{\protect\hyperlink{ref-gill2022}{54}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{58}},
suggest that the type of occupation in individuals of equity-deserving
groups played an important role in increasing the odds of vaccination.

However, the results also indicate that the place of habitation affected
the odds of vaccination for certain equity-deserving groups (interaction
term of Health Region and Race, Figure~\ref{fig-models},B).
Specifically, this held true in the case of individuals identifying as
Indigenous or with mixed racial background in the West Health Region,
Black individuals in the Central Health Region, and Arab/Middle Eastern
individuals in the East Health Region Figure~\ref{fig-models}. For these
individuals, vaccination odds were lower when compared to the Toronto
Health Region (Supplementary Figure A-6). We indicate next some
contributing factors that might help provide context to these results.

>>>>

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata},
  pdfauthor={Ariel Mundo Ortiz; Bouchra Nasri},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Differences in COVID-19 vaccination in the province of
Ontario across Health Regions and socio-economic strata}}

\author[1,2,3]{Ariel Mundo Ortiz}
\author[1,2,3,*]{Bouchra Nasri}

\affil[1]{Centre de Recherches Mathématiques, University of Montreal,
Montréal, Canada}
\affil[2]{Department of Social and Preventive Medicine, École de Santé
Publique, University of Montreal, Montréal, Canada}
\affil[3]{Centre de recherche en santé publique, University of Montreal,
Montréal, Canada}
\affil[*]{Corresponding author, \url{bouchra.nasri@umontreal.ca}}

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, interior hidden]}{\end{tcolorbox}}\fi

@@

As of May of 2023, there have been 765 million confirmed cases of
COVID-19 around the world, including 6.8 million
deaths\textsuperscript{\protect\hyperlink{ref-WHO-Covid}{1}}. Although
this disease is no longer categorized as a global health emergency by
the World Health Organization
(WHO)\textsuperscript{\protect\hyperlink{ref-rigby2023}{2}}, there is
ongoing concern due to continued transmission, surges in cases and
deaths due to new
variants\textsuperscript{\protect\hyperlink{ref-un2023}{3}}, and
weaknesses in health systems around the world that could be exploited by
a novel virus or another public health emergency in the
future\textsuperscript{\protect\hyperlink{ref-mackey2021}{4}}.

In particular, a major weakness that has received attention during the
pandemic has been related to inequalities in vaccine uptake. The rapid
development of vaccines against COVID-19 initially brought the hope of a
rapid end to the pandemic due to the start of vaccination campaigns in
certain parts of the world toward the end of
2020\textsuperscript{\protect\hyperlink{ref-thelancet2021}{5}--\protect\hyperlink{ref-tanne2020}{8}})
but inequalities in vaccine uptake made these pharmaceutical
interventions ultimately unable to replicate the success of the smallpox
vaccination program, which was crucial to control this
disease\textsuperscript{\protect\hyperlink{ref-kayser2021}{9}}.

This problematic is a multifaceted issue resulting from a combination of
factors, among which are failed public health
measures\textsuperscript{\protect\hyperlink{ref-li2021}{10}}, inequality
in vaccine access between high- and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{11},\protect\hyperlink{ref-tamey2022}{12}},
and vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{13}}.
Furthermore, it is well established that this issue has affected in
particular individuals in certain underrepresented groups (e.g., Black,
Asian, or Indigenous) as well as individuals with socio-economic
disadvantages\textsuperscript{\protect\hyperlink{ref-willis2021}{14}--\protect\hyperlink{ref-hussain2022}{20}}.

Reasons given for this inequality have included medical mistrust due to
systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{16},\protect\hyperlink{ref-mosby2021}{21}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{14}}, and
the influence of conspiracy
theories\textsuperscript{\protect\hyperlink{ref-mosby2021}{21}--\protect\hyperlink{ref-freeman2020}{23}}.
However, it is important also to consider that vaccination uptake can be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{24}},
spatial differences in vaccine access and supply, vaccination location
availability, and lack of prioritization of areas where vulnerable
groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-nguyen2021}{25}}.
Other studies have also shown heterogeneity in vaccine uptake within
small governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{26}--\protect\hyperlink{ref-bhuiyan2022}{29}},
and that accounting for geographical differences in vaccination can help
predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{30}}.

However, such analyses have been carried out mostly in territories
outside of Canada, where available studies have been focused in certain
cities (such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{31}}, or
Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{32}}), or
have explored differences at a province-wide
level\textsuperscript{\protect\hyperlink{ref-guay2022}{18}}. Therefore,
there is a need for studies that explore spatial differences in
vaccination within the Canadian territory and that consequently, can
help identify disparities that need to be addressed within specific
areas in each province.

This need is especially important in the case of Ontario, the most
populated province of Canada. Between 2007 and 2019, Ontario managed
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integration Networks (LHINs), which aimed to
provide an integrated health system for the province. However, this
approach was complex, bureaucratic, and resulted in excessive
expenditures, disparities in mortality rates, the deterioration of
certain performance indicators such as wait times and hospital
readmissions, fragmented electronic health systems, the decline of
performance indicators, and inequities in health services
access\textsuperscript{\protect\hyperlink{ref-tsasis2012}{33}--\protect\hyperlink{ref-lysyk2016}{37}}.
Therefore, with the intent of better organizing and delivering care in
late 2019 the provincial government eliminated the LHINs and
incorporated the areas covered by them into six larger Health Regions
(North East, North West, Central, Toronto, West, and
East)\textsuperscript{\protect\hyperlink{ref-dong2022}{35}}.

Because of the relatively recent adoption of the Health Region model and
its alignment with the onset of the COVID-19 pandemic, there is a need
to analyze if there are ongoing disparities in health access under this
approach that need to be addressed before they are exploited by a new
disease or public health threat. In this regard, previous research has
highlighted disparities in the level of activity of each Health
Region\textsuperscript{\protect\hyperlink{ref-sethuram2023}{38}}.
Therefore, analyzing differences in vaccination uptake within the Health
Regions and can help identify which socio-demographic groups are the
most vulnerable and what areas of the province deserve special attention
by decision-makers.

Therefore, in this study we aim to understand the differences in
vaccination uptake between the different Health Regions of Ontario
between October of 2021 and January of 2022. By including socio-economic
factors in our analysis, we aimed to identify in which groups these
differences were significant in order to provide an assessment of the
current state of healthcare access in Ontario.

@@

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences and the Mathematical Modelling of
COVID-19 Task Force under ethical guidance from the University of
Toronto (under protocol 00043317), and which ran between September 30th,
2021 and January 17th,2022. The survey collected socio-economic
information from participants (Table~\ref{tbl-descriptive-stats}), their
location (nearest municipality, as shown in Figure~\ref{fig-map}), the
date of access to the survey, and asked information on vaccination
status by using the question ``Have you received the first dose of the
COVID vaccine?'', with possible answers ``yes'' and ``no''. The original
dataset contained 39,029 observations.

By design, the survey allowed respondents to exit at any time and
deployed the questions randomly, which resulted in \(\approx\) 84\% of
the observations having multiple missing answers or being incomplete.
Therefore, we selected 6,343 observations that were labelled as
``complete'' in the dataset and that had answers for all covariates
considered in our analysis. Later, we matched the city of each
observation with its corresponding LHIN and Health Region, and removed
observations from areas with low representation (254 observations
corresponding to the North West and North East Health Regions). Finally,
we removed outliers from the data (19 observations of individuals under
25 years of age with household size of 1 and income above 110,000 CAD in
the original dataset, that were demeed to lack accuracy). After all the
preliminary analyses, the total number of observations used for analysis
was 6,236 and included the East, Central, Toronto, and West Health
Regions covering between October 1st,2021 and January 17, 2022. The
original dataset, clean dataset, and details on the data cleaning
process are described in detail in the
\href{https://github.com/aimundo/Fields_COVID-19/}{GitHub repository}
for this paper.

@@

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{39}--\protect\hyperlink{ref-cnat2022a}{41}}.
Because we identified differences in representativity between the survey
data and the estimates from the Census, we used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{42}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{43}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{44}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{45}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{46}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{47}}.

@@

Figure~\ref{fig-models} presents the estimates (as odd ratios) from the
logistic regression models for vaccination status using the
socio-demographic factors collected by the survey, and their
interactions. Generally speaking, lower odds of vaccination were
identified in both cases in individuals characterized by a low household
income, or that identified as part of underrepresented groups. However,
the magnitude of the estimates differed between the uncorrected and
corrected models and more importantly, there were differences in the
statistical significance of certain estimates before and after the
correction. Specifically, the uncorrected model showed significant
differences in vaccination odds between the age groups considered, the
East Health Region, Latin American individuals with a household income
under CAD 25,000, and Indigenous individuals living in the Central
Health Region (Figure~\ref{fig-models},B) but these were deemed non
statistically significant after the correction.

@@

Interestingly, individuals in underrepresented groups with a household
income below CAD 25,000 had higher odds of vaccination (when compared to
those with a household income above CAD 60,000). This held true in the
case of Arab/Middle Eastern individuals (OR=34, CI={[}1.70,6.79{]}),
Black individuals (OR=3.81, CI={[}2.05, 7.09{]}), and those in other
racial or ethnic groups (OR=3.19, CI={[}2.00,5.09{]}). Additionally,
individuals with an income between CAD 25,000 and 59,999 in the
Arab/Middle Eastern and other racial or ethnic groups also had higher
odds of vaccination than their high-income peers (ORs and CIs=6.96
{[}2.67,18.16{]}, and 3.5 {[}1.85,6.62{]}).

Finally, \textbf{the place of habitation affected the odds of
vaccination for certain underrepresented groups, as significantly lower
odds of vaccination were identified for the interaction between Health
Region and race} in the case of Black individuals in the Central Health
Region (OR=0.39, CI={[}0.2,0.75{]}), Arab/Middle Eastern individuals in
the East Health Region (OR=0.41 {[}0.17, 0.98{]}), and in the Indigenous
and mixed groups in the West Health Region (ORs and CIs={[}0.31 {[}0.14,
0.7{]} and 0.38 {[}0.19, 0.76{]}, respectively).

@@

In this study, we hypothesized that differences in COVID-19 vaccination
uptake were present between the Health Regions between late 2021 and
early 2022. Our goal was to determine which socio-demographic groups
could be impacted by these disparities in order to provide
decision-makers with information that could be used to develop policies
focused on reducing or eliminating these differences and ensure that the
Health Region model is able to fulfill its mission of improving health
access for all Ontarians. \textbf{making corrections up to this point}

@@

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in underrepresented Canadian individuals with concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{41},\protect\hyperlink{ref-basta2022}{49}--\protect\hyperlink{ref-cnat2023}{51}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of underrepresented
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{52}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{53}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with vaccine
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{54}}. This
suggests that the observed differences are associated with disparities
in vaccine access that were present during the period covered by the
survey.

Interestingly, whereas overall self-reported vaccination rates were
found to be statistically significantly lower in various
underrepresented groups when compared to White/Caucasian individuals,
the change in odds of vaccination within certain racial groups and
income strata was actually positive, in contrast to the White/Caucasian
group, where vaccination odds decreased in income brackets below CAD
60,000 (Supplementary Figure A-5). Specifically, individuals in low
income brackets that belonged to Arab/Middle Eastern, Black, or other
minority groups had higher odds of vaccination that their peers with an
income above 60,000 CAD.

This result is likely reflects in part the fact that individuals in
underrepresented groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{55},\protect\hyperlink{ref-ct2021}{56}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. In Ontario, these workers had priority for
COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{57}}; and
there is evidence of interventions by vaccination staff in certain parts
of the province to encourage vaccination uptake by these
individuals\textsuperscript{\protect\hyperlink{ref-gill2022}{54}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{58}},
suggest that the type of occupation of individuals in underrepresented
groups played an important role in increasing the odds of vaccination.

However, the results also indicate that the place of habitation affected
the odds of vaccination for certain underrepresented groups (interaction
term of Health Region and Race, Figure~\ref{fig-models},B).
Specifically, this held true in the case of individuals identifying as
Indigenous or with mixed racial background in the West Health Region,
Black individuals in the Central Health Region, and Arab/Middle Eastern
individuals in the East Health Region Figure~\ref{fig-models}. For these
individuals, vaccination odds were lower when compared to the Toronto
Health Region (Supplementary Figure A-6). We indicate next some
contributing factors that might help provide context to these results.
